---
title: "Formatting popler datasets"
author: "Aldo Compagnoni"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to popler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Start from the metadata

To pick the studies you want to use, start looking at the metadata. Metadata information is obtained from the `browse` function. Browse contains all the metadata from all of the studies currently in popler. Running `browse` simply returns the metadata of all the studies currently in popler:

```{r, message = FALSE, warning = FALSE}
#devtools::install_github("AldoCompagnoni/popler", build_vignettes = TRUE)
library(popler)
browse()
```

To zoom in the type of studies you are most interested in, you can define the criteria in the browse function. Say that you'd like only observational studies, with community data (data on more than 1 species), more than 20 years long, and with more than 5 spatial replicates:

```{r, message = FALSE, warning = FALSE}
browse(duration_years>20 & studytype == 'obs' & community == 'yes' & lterid != 'SBC' &         tot_spat_rep > 5 )
```

Above, you can see the study `title`, the ID of each study (`proj_metadata_key`), the three letter of the LTER site where the study was conducted (`lterid`), the type of abundance data contained in the dataset (`datatype`), whether the study is observational or experimental (`studytype`, potential values are 'obs' or 'exp'), and duration in years (`duration_years`). This file is much larger, but this shall suffice.

## How to see if a study is what you are looking for

Before you download a data frame to format it, you 1) find it and 2) verify that it is actually what you are looking for. The easiest way to do this is to run browse by adding an argument `report = T`. This argument will open up a legible html document.
This document produces open html page in your browser which have hyperlinks, and a list of the studies you selected through browse. Use this document to look up the description of each study, and, most importantly, look at each study's its ORIGINAL METADATA. To look at the original metadata, click on the hyperlink named `metadata_link`. 

```
browse(duration_years>20 & studytype == 'obs' & community == 'yes' & lterid != 'SBC' & tot_spat_rep > 5 , report = T)

```

Now, say that I checked the studies title, description, and, `metadata_link`. I ended up determining that the study called "SGS-LTER Standard Production Data : 1983-2008..." is what I am looking for. In popler, this study has ID number 65. To download this study, simply 

```{r, message = FALSE, warning = FALSE}
# download SGS biomass data from popler
sgs_biom_raw <- get_data(proj_metadata_key == 65)
```

### Format "Taxon counts"

First, format the actual abundance information from the original data set. Below is a snipped of code that performs that. I paste the code below, and then I explain its features.

```
sgs_biom_ab  <- sgs_biom_raw %>%
                  mutate( OBSERVATION_TYPE = 'TAXON_COUNT',
                          # Site id is a combination of the three nested spatial levels
                          SITE_ID          = paste(spatial_replication_level_1, 
                                                   spatial_replication_level_2, 
                                                   spatial_replication_level_3, sep = "_"),
                          DATE             = year,
                          # in this case X2_value == 'Species name"
                          VARIABLE_NAME    = paste(genus, species, sep=" "),
                          VARIABLE_UNITS   = NA,
                          VALUE            = abundance_observation) %>% 
                  select( OBSERVATION_TYPE, SITE_ID, DATE, VARIABLE_NAME, VARIABLE_UNITS, VALUE)
```

OBSERVATION_TYPE states that here we refer to "TAXON_COUNT" (the abundance of each species).

SITE_ID shows each separate spatial replicate. Most LTER studies are *spacially nested*. Not so in the *metacomm* working group format. Hence, you have to consider every combination of spatial replicates as a separate spatial replicate. To do so, 
  1. identify every column starting with `spatial_replication_level_`
  2. paste these columns using the `paste` function and argument `sep="_"`
The dataset I have downloaded here has three nested spatial levels. 

DATE simply refers to the year. Every popler dataset provides a column with year, so this should not be an issue. UNLESS, censuses in the study were more frequent than once a year.

VARIABLE_NAME in taxon count data is the species name. In this case, we are lucky enough to have both `genus` and `species` columns. Simply join them together using the `paste` function.
Sometimes, however, you will only have the species code data, contained in a column called `sppcode`. In this case, simply assign `sppcode` to VARIABLE_NAME.

VARIABLE_UNITS, and VALUES: you should always run the last three lines of code above!


### Format LATITUDE/LONGITUDE information using the cov_unpack function

IMPORTANT NOTICE: few popler datasets have a latitude and longitude information for all spatial replicates. We sometimes provide lat/lon information associated with the  `spatial_replication_level_1` (the site). 
However, some popler datasets have latitude and longitude information in the `covariates` column. This is a column that "packs" all information that does not fit into the structure of the popler database. In this example, the covariates contain latitude and longitude information:

```{r, message = FALSE, warning = FALSE}
# download SGS biomass data from popler
sgs_biom_raw$covariates  %>% head
```

You can unpack this column of gibberish into a proper dataframe using function `cov_unpack`:

```{r, message = FALSE, warning = FALSE}
# download SGS biomass data from popler
sgs_biom_raw  %>% cov_unpack(.) %>% head
```

In this dataframe, X1_label shows what each value refers to. Hence, `X1_label` tells us that `X1_value` contains "comments". `X2_label` tells us that `X2_value` contains a "scientific name" and so on.

Given the above, now we can format two separate data farames: one containing latitude information, the other containing longitude information. NOTE the `fac_char_num` function - I wrote this to convert factor variables into a numeric value.

```
# Spatial coordinates

# function to convert lat/lon from factor, to character, to numeric
fac_char_num <- function(x) x %>% as.character %>% as.numeric

# Format LATITUDE
sgs_biom_lat <- sgs_biom_raw %>%
                  cbind( cov_unpack(.) ) %>% 
                  mutate( OBSERVATION_TYPE = 'SPATIAL_COORDINATE',
                          SITE_ID          = paste(spatial_replication_level_1, 
                                                   spatial_replication_level_2, 
                                                   spatial_replication_level_3, sep = "_"),
                          DATE             = NA,
                          # in this case X2_value == 'Species name"
                          VARIABLE_NAME    = 'LATITUDE',
                          VARIABLE_UNITS   = 'decimal.degrees',
                          VALUE            = fac_char_num(X4_value) ) %>% 
                  select( OBSERVATION_TYPE, SITE_ID, DATE, 
                          VARIABLE_NAME, VARIABLE_UNITS, VALUE) %>% 
                  unique

# Format LONGITUDE
sgs_biom_lon <- sgs_biom_raw %>%
                  cbind( cov_unpack(.) ) %>% 
                  mutate( OBSERVATION_TYPE = 'SPATIAL_COORDINATE',
                          SITE_ID          = paste(spatial_replication_level_1, 
                                                   spatial_replication_level_2, 
                                                   spatial_replication_level_3, sep = "_"),
                          DATE             = NA,
                          # in this case X2_value == 'Species name"
                          VARIABLE_NAME    = 'LONGITUDE',
                          VARIABLE_UNITS   = 'decimal.degrees',
                          VALUE            = fac_char_num(X5_value) ) %>% 
                  select( OBSERVATION_TYPE, SITE_ID, DATE, 
                          VARIABLE_NAME, VARIABLE_UNITS, VALUE) %>% 
                  unique
```

## Put it all together

Finally, you can stack these three files file together with a slick one line of code:

```
SGS_BIOMASS <- Reduce(function(...) rbind(...), list(sgs_biom_lat,sgs_biom_lon,sgs_biom_lat) )
```
